PeakNet: An Autonomous Bragg Peak Finder with Deep Neural Networks

### 2023-06-16 manuscript.tex
Correct typos.

Serial crystallography with X-ray free electron lasers (XFEL) has enabled the structural determination of three-dimensional macromolecules from radiation damage-free samples at room temperature, an approach commonly known as \textit{diffraction-before-destruction} \citep{neutzePotentialBiomolecularImaging2000, chapmanFemtosecondDiffractiveImaging2006,chapmanFemtosecondXrayProtein2011}.  Time-resolved serial crystallography at XFEL facilities, like the Linac Coherent Light Source (LCLS), allows the investigation of biochemical reactions with femtosecond-scale time resolutions.  This technique is typically referred to as serial femtosecond crystallography (SFX), which has led to many signficiant structural studies \citep{kupitzSerialTimeresolvedCrystallography2014, nangoThreedimensionalMovieStructural2016,pandeFemtosecondStructuralDynamics2016a, youngStructurePhotosystemII2016,sugaLightinducedStructuralChanges2017, kernStructuresIntermediatesKok2018,ibrahimUntanglingSequenceEvents2020, sugaTimeresolvedStudiesMetalloproteins2020} since the first published work at LCLS \citep{aquilaTimeresolvedProteinNanocrystallography2012}.  In the meantime, the data rate at XFEL facilities have also gradually increased from hundreds of Hz to the MHz range, with the European X-ray Free Electron Laser (EuXFEL) being the first MHz facility available to users and LCLS-II set to launch in 2023.

@@

While template matching and region-growing tecnhiques require users to provide priori information like threshold values, Robust Peak Finder (RPF) \citep{hadian-jaziPeakfindingAlgorithmBased2017, hadian-jaziDataReductionSerial2021} has demonstrated proficient peak finding capabilities with minimal user inputs.  This techinuqe utilizes the Modified Selective Statistical Estimator (MSSE) method to segment between background pixels and actual peaks.  Furthermore, the RPF method can also parallelize the processing of diffraction patterns, making it ideal for real-time data analysis.

\psocake{} is a software specifically designed to streamline high throughput data reduction and analysis of XFEL experiments, and is routinely used at SLAC National Accelerator laborator (SLAC) and Pohang Accelerator Laboratory (PAL).  The built-in peak-finding algorithm \citep{shinDataAnalysisUsing2018} includes calibrating raw analog digital units (ADUs); applying optional background correction and bad pixel masks; identifying possible peaks; calculating their SNR; determining Bragg spot sizes; and selecting peaks based on size, total intensity and SNR.

@@

where CFL stands for categorical focal loss, $\alpha_j$ is a balancing factor for each class, $y_j^{(i)}$ is the grouth truth label of the $j$-th class for the $i$-th example.  $p_j^{(i)}$ is the predicted probability of the $j$-th class for the $i$-th example, calculated by the neural network $f$ parameterized by $\theta$ given input $x_j^{(i)}$.  $\gamma$ is a parameter controling the extent to which the weight of well-classified examples, such as background pixels, is reduced.  In our neural network training, we chose $\gamma = 2.0$, essentially, this means $(1-\hat{p} _j^{(i)}) ^\gamma$ is solely responsible for down-weighting easy examples, as $\hat{p} _j^{(i)}$ is usually large in this case.  Finding the ideal values for $\alpha$ can be a delicate process with only marginal returns, as highlighted in the original focal loss study \citep{linFocalLossDense2018}.  In order to fully utilize focal loss, we decided to use a fixed trivial value of $\alpha = 1.2$ for all examples, primarily for the sake of completeness.  This choice essentially treats $\alpha$ as a simple scaling factor without significant impact.  It is worth noting that in most cases, the sum of all $\alpha_j$ values, denoted as $\sum_{j = 1}^{C} \alpha_j$, is expected to equal 1.

@@

However, for the proteinase (PK) sample, the performance comparision is mixed.  Although \peaknet{} lags slightly behind manually fine-tuned \psocake{} in terms of R-split and redundancy, it outperforms in CC$_1/2$, CC*, $I/\sigma(I)$, and completeness values.  This reinforces the effectiveness of \peaknet{} in achieving respectable results across varied sample types, without any manual intervention or parameter fine-tuning.

Overall, \peaknet{} demonstrates comparable performance to a manually fine-tuned \psocake{} for two distinct samples, showcasing its potential for autonomous peak finding with consistent results.  By reducing dependence on the expertise of individuals, which can have considerable variation, \peaknet{} offers a promising user-indepenent solution.  Notably, despite primarily being trained on PK samples, \peaknet{} maintains robust and versatile performance, indicating its broad applicability and effectiveness in handling diverse datasets.

@@

\peaknet{} exhibits exceptional capabilities in autonomous Bragg peaks finding without requiring user input or intervention, such as algorithmic parameter tuning or manual artifact scattering masking.  The predictive performance of \peaknet{} in peak finding proves to be comparable to, if not better than, manually fine-tuned \psocake{} with user-supplied masks.  An integral feature of \peaknet{}, auto-masking, significantly enhances its ability to autonomously identify and process artifact scattering patterns.  Moreover, the efficiency of \peaknet{} is exemplified by its impressive runtime performance, surpassesing commonly used peak finders like \psocake{} and RPF.  Parallel processing with GPU streaming can potentially improve the runtime efficiency even further.  In conclusion, \peaknet{}  offers a robust, versatile, and efficient solution that operates independently of user input, thereby holding significant potential for optimizing crystallographic data processing at high data rates.

>>>>

Serial crystallography with X-ray free electron lasers (XFEL) has enabled the structural determination of three-dimensional macromolecules from radiation damage-free samples at room temperature, an approach commonly known as \textit{diffraction-before-destruction} \citep{neutzePotentialBiomolecularImaging2000, chapmanFemtosecondDiffractiveImaging2006,chapmanFemtosecondXrayProtein2011}.  Time-resolved serial crystallography at XFEL facilities, like the Linac Coherent Light Source (LCLS), allows the investigation of biochemical reactions with femtosecond-scale time resolutions.  This technique is typically referred to as serial femtosecond crystallography (SFX), which has led to many significant structural studies \citep{kupitzSerialTimeresolvedCrystallography2014, nangoThreedimensionalMovieStructural2016,pandeFemtosecondStructuralDynamics2016a, youngStructurePhotosystemII2016,sugaLightinducedStructuralChanges2017, kernStructuresIntermediatesKok2018,ibrahimUntanglingSequenceEvents2020, sugaTimeresolvedStudiesMetalloproteins2020} since the first published work at LCLS \citep{aquilaTimeresolvedProteinNanocrystallography2012}.  In the meantime, the data rate at XFEL facilities have also gradually increased from hundreds of Hz to the MHz range, with the European X-ray Free Electron Laser (EuXFEL) being the first MHz facility available to users and LCLS-II set to launch in 2023.

@@

While template matching and region-growing techniques require users to provide priori information like threshold values, Robust Peak Finder (RPF) \citep{hadian-jaziPeakfindingAlgorithmBased2017, hadian-jaziDataReductionSerial2021} has demonstrated proficient peak finding capabilities with minimal user inputs.  This technique utilizes the Modified Selective Statistical Estimator (MSSE) method to segment between background pixels and actual peaks.  Furthermore, the RPF method can also parallelize the processing of diffraction patterns, making it ideal for real-time data analysis.

\psocake{} is a software specifically designed to streamline high throughput data reduction and analysis of XFEL experiments, and is routinely used at SLAC National Accelerator laboratory (SLAC) and Pohang Accelerator Laboratory (PAL).  The built-in peak-finding algorithm \citep{shinDataAnalysisUsing2018} includes calibrating raw analog digital units (ADUs); applying optional background correction and bad pixel masks; identifying possible peaks; calculating their SNR; determining Bragg spot sizes; and selecting peaks based on size, total intensity and SNR.

@@

where CFL stands for categorical focal loss, $\alpha_j$ is a balancing factor for each class, $y_j^{(i)}$ is the grouth truth label of the $j$-th class for the $i$-th example.  $p_j^{(i)}$ is the predicted probability of the $j$-th class for the $i$-th example, calculated by the neural network $f$ parameterized by $\theta$ given input $x_j^{(i)}$.  $\gamma$ is a parameter controlling the extent to which the weight of well-classified examples, such as background pixels, is reduced.  In our neural network training, we chose $\gamma = 2.0$, essentially, this means $(1-\hat{p} _j^{(i)}) ^\gamma$ is solely responsible for down-weighting easy examples, as $\hat{p} _j^{(i)}$ is usually large in this case.  Finding the ideal values for $\alpha$ can be a delicate process with only marginal returns, as highlighted in the original focal loss study \citep{linFocalLossDense2018}.  In order to fully utilize focal loss, we decided to use a fixed trivial value of $\alpha = 1.2$ for all examples, primarily for the sake of completeness.  This choice essentially treats $\alpha$ as a simple scaling factor without significant impact.  It is worth noting that in most cases, the sum of all $\alpha_j$ values, denoted as $\sum_{j = 1}^{C} \alpha_j$, is expected to equal 1.

@@

However, for the proteinase (PK) sample, the performance comparison is mixed.  Although \peaknet{} lags slightly behind manually fine-tuned \psocake{} in terms of R-split and redundancy, it outperforms in CC$_1/2$, CC*, $I/\sigma(I)$, and completeness values.  This reinforces the effectiveness of \peaknet{} in achieving respectable results across varied sample types, without any manual intervention or parameter fine-tuning.

Overall, \peaknet{} demonstrates comparable performance to a manually fine-tuned \psocake{} for two distinct samples, showcasing its potential for autonomous peak finding with consistent results.  By reducing dependence on the expertise of individuals, which can have considerable variation, \peaknet{} offers a promising user-independent solution.  Notably, despite primarily being trained on PK samples, \peaknet{} maintains robust and versatile performance, indicating its broad applicability and effectiveness in handling diverse datasets.

@@

\peaknet{} exhibits exceptional capabilities in autonomous Bragg peaks finding without requiring user input or intervention, such as algorithmic parameter tuning or manual artifact scattering masking.  The predictive performance of \peaknet{} in peak finding proves to be comparable to, if not better than, manually fine-tuned \psocake{} with user-supplied masks.  An integral feature of \peaknet{}, auto-masking, significantly enhances its ability to autonomously identify and process artifact scattering patterns.  Moreover, the efficiency of \peaknet{} is exemplified by its impressive runtime performance, surpassing commonly used peak finders like \psocake{} and RPF.  Parallel processing with GPU streaming can potentially improve the runtime efficiency even further.  In conclusion, \peaknet{}  offers a robust, versatile, and efficient solution that operates independently of user input, thereby holding significant potential for optimizing crystallographic data processing at high data rates.
